TacticToe: Learning the Game of Tactics

### 2018-01-24 tactictoe_jv.tex
related works: wip
updating tactictoe data with the 60 seconds version

\subsection{The problem}

\todo{JU or CK review}
\subsection{Contributions}
This project is an extension of the work described in \cite{tgckju-lpar17}. We 
explain here the additional contributions and their effect across our system:

@@

%Mention explictly all the predictors that will be using this distance.
%tactics:
%theorems:
%terms:
%goal list: (product of tactics)
\begin{itemize}
\item tactic:\\
Given a goal $g$, the classifier selects a list of previously solved goals $l$
ordered by their similarity to $g$. Since each solved goal is associated 
with a tactic in our database, the ordered list of goals induces a ordered  
list of associated tactics. And tactics are tried in this order during the 
proof search. Therefore, to order the tactics correctly, it is crucial to 
estimate the similarity between goals with mathematically relevant features.
\item term:\\

\item theorem:\\

\item list of goals:\\

\end{itemize}

@@

%However for very large branching factor RTS Combinatorial 
%Multi-armed Bandits
%for Real-Time Strategy Games. The method of choice for solving those 
%problems 
%is monte carlo tree search~\cite{mcts} which is a way to assign probabilities 
%to each branches

So, we first give a static description of the proof tree and its component.
Then we explain how we compute the evaluation and policy functions from learned 
data. And finally, we show how the two MCTS step, node selection and node 
extension changes the proof tree data structures.

%A proof is in general not a path but a tree. This means that our search space 
%has two branching factors: 
%the choice of a tactic, and the number of goals produced by each tactic.
%The first one can be thought as a disjunction of tactics. Indeed, 
%if there exists a tactic that solves a goal then it is solved. And 
%conversely, the second branching factor is a conjunction of goals. All pending 
%goals must be proven in order to prove the parent goal. For example, in the 
%case of induction tactics the pending goals are the base case and the 
%inductive 
%case.

%If we suppose that each tactic produces at most one goal, then the standard 
%monte carlo tree search algorithm applies. So we will first discuss how our 
%algorithm works in this case. And then explain how we extend it to support 
%tactics producing 
%multiple goals.

%Our search algorithm takes inspiration from the A*-algorithm~\cite{HartNR68} 
%which uses 
%a cost function and heuristics to estimate the total distance to the 
%destination and
%choose the shortest route.
%The proof is not finished when the current tactic solves its goal because it 
%often leaves new pending open goals along the current path.

%\subsection{Diagonalized tree search}
%(Move this section to Monte Carlo tree search and simplify it) 
%(Don't mention diagonlazied tree search)

%%
%%Our first attempt was to create a breadth first search algorithm with 
%%preferences for the best predicted tactic for each goal. In 
%%Figure~\ref{fig:br1}, we try to prove the goal $g$ using list of tactics 
%%$t_1,t_2,t_3 \ldots$ ordered by the predictor. Those tactics produce 
%%respectively the goals $g_1,g_2,g_3 \ldots$. By choosing tactics
%%for each produced goal $g_i$, we can recursively construct the whole search 
%%tree. We observe that if we apply the breadth first search algorithm to this 
%%tree then the predictor preferences are of little use, as we will not explore 
%%deeper more promising branches. Another related inconvenience is that we may 
%%not ever be able to explore the tree at depth 2 since there maybe too many 
%%tactics to try at depth 1. One way to solve those this problem is to 
%%diagonalized our tree by adding dummy square nodes as presented in 
%%Figure~\ref{fig:diag_choice}. The new branching factor of this tree is 
%%constant 
%%equal to two. The exploration of lower ranked tactics are delayed by the 
%%diagonalization during breadth search. And after $2^n$ breadth-first steps, 
%%the 
%%most promising branch has been explored up to depth $n$.
%%
%%\begin{figure}{}
%%\begin{center}
%%\begin{tikzpicture}[node distance=2cm]
%%\node [tcircle] (0) {$g$};
%%\node [above of=0] (m0) {};
%%\node [tcircle, left of=m0, node distance=1cm] (2) {$g_2$};
%%\node [tcircle, left of=2] (1) {$g_1$};
%%\node [tcircle, right of=m0,node distance=1cm] (3) {$g_3$};
%%\node [right of=3] (inf) {};
%%
%%\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1);
%%\draw[-to,thick] (0) to node[xshift=-10](t2){$t_2$} (2);
%%\draw[-to,thick] (0) to node[xshift=-10](t3){$t_3$} (3);
%%\draw[-to,dotted,thick] (0) to (inf);
%%\end{tikzpicture}
%%%\end{center}
%%%\caption{Choice of a tactic}
%%%\label{fig:choice}
%%%\end{figure}
%%%\begin{figure}{}
%%%\begin{center}
%%\begin{tikzpicture}[node distance=1cm]
%%\node [tcircle] (0) {$g$};
%%\node [above of=0] (m0) {};
%%\node [tcircle, left of=m0] (1) {$g_1$};
%%\node [tsquare, right of=m0] (s1) {};
%%\node [above of=s1] (m1) {};
%%\node [tcircle, left of=m1] (2) {$g_2$};
%%\node [tsquare, right of=m1] (s2) {};
%%\node [above of=s2] (m2) {};
%%\node [tcircle, left of=m2] (3) {$g_3$};
%%\node [right of=m2] (s3) {};
%%
%%\draw[-to,thick] (0) to node[xshift=-5,yshift=-5](t1){$t_1$} (1);
%%\draw[-to,thick] (0) to (s1);
%%\draw[-to,thick] (s1) to node[xshift=-5,yshift=-5](t1){$t_2$} (2);
%%\draw[-to,thick] (s1) to (s2);
%%\draw[-to,thick] (s2) to node[xshift=-5,yshift=-5](t1){$t_3$} (3);
%%\draw[-to,dotted,thick] (s2) to (s3);
%%\end{tikzpicture}
%%\end{center}
%%\caption{Transformation of a search tree into a diagonalized search tree,
%%caused   by a change in the relation between a node and its 
%%children.}
%%\label{fig:diag_choice}
%%\end{figure}
%%
%%
%%In the particular case where tactics produces at most one goal, a proof is 
%%found in the search tree when a tactic $t$ does not produce any new goal. The 
%%proof of the root goal can be reconstructed by extracting the path leading 
%%to the final tactic $t$.
%%
%%
%%
%%\subsection{General case}
%%We now tackle the more general case where tactics $t_i$ may produces multiple 
%%goals $g_{i,1}$ $g_{i,2}$ $g_{i,n}$ from a parent goal $g_p$.
%%
%%All created node are kept in memory for later inspection but some of the 
%%nodes 
%%are deactivated which means they are removed from the proof search.
%%
%%Definition (deactivated node):
%%  A solved node or a node that has a solved ancestor.
%%
%%Then at first, the breadth first search operates normally with only the first 
%%goal $g_{i,1}$ is present in the search tree. 
%%And as soon as a proof is found for the goal $g_{i,1}$, the subtree with root 
%%$g_{i,1}$ is erased and replaced by a node containing $g_{i,2}$. The process 
%%continues with node with the lowest depth being treated first.
%%The process stops if the root goal is solved, the time limit is exceeded or 
%%there is no more tactic to be applied to an active node.

@@

\begin{definition}(Tactical search tree)\\
A tactical search tree $\mathfrak{T}$ is a sixtuple 
$(\mathbb{A},\mathbb{G},\mathbb{T},A,G,T)$ 
that respects the following conditions:
\begin{itemize}
\item $\mathbb{A}$ is a set of nodes, $\mathbb{G}$ is a set of goals 
 and $\mathbb{T}$ is a set of tactics.
\item $A$ is a partial function from $\mathbb{G} \times \mathbb{T}$ to 
$\mathbb{A}$.
\item $T$ is a total function from $\mathbb{G}$ to $P(\mathbb{T})$.
\item $G$ is a total function from $\mathbb{A}$ to $P(\mathbb{G})$.
\item $Support(A) = \lbrace (g,t)\ |\ t \in T(g) \rbrace$.
\item if $t \in T(g)$ and $A(g,t) = a$ then $t(g) = G(a)$.
\end{itemize}

@@

In each node can either be solved, open or pending. The definition of a solved 
goal is expressed through mutual recursion with a 
solved node and is given in ~\ref{def:solved_goal}. An open goal is goal that 
was tried at least once by a tactic. In our implementation we have only one 
open goals per node at once. A goal that was not explored yet is called a 
pending goal.

@@

%A proof is in general not a path but a tree. This means that our search space 
%has two branching factors: 
%the choice of a tactic, and the number of goals produced by each tactic.
%The first one can be thought as a disjunction of tactics. Indeed, 
%if there exists a tactic that solves a goal then it is solved. And 
%conversely, the second branching factor is a conjunction of goals. All pending 
%goals must be proven in order to prove the parent goal. For example, in the 
%case of induction tactics the pending goals are the base case and the 
%inductive 
%case.

@@

Since the support of A induces a bijection between tactics of the tree 
search and the node they 
produced, any function define on tactics of the tree search can be map to a 
function defined on nodes. 
Therefore, we define $PriorPolicy(a)=_def PriorPolicy(t)$, if $\exists g 
A(g,t) = a$.

@@

\subsection{Evaluation}

@@

Mention the conflicts.
Theorem orthogonalization is not turn on during instantiation of tacitcs except 
for the metis tactic.
Best parameters + comparison with holyhammer.
asynchronous eprover is not used  for fairer comparison with eprover.

10 seconds experiment:

Until now, all our feature vectors were learned form human proofs. 
human proofs. We can also add to this tactics that were part of a proof found 
by our algorithm. To prevent duplication of effort, orthogonalization of those 
tactics is essential to have a beneficial effect (see Experiments).
Since recording and re-proving are intertwined, the 
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own 
discovered proofs than from the human proof 
scripts~\cite{DBLP:conf/cade/Urban07}. It is also the first step toward 
self-reinforcement learning.

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{llll}
\toprule
  & & Small(273) & Big(7088) \\
\midrule
  & \eprover & 86 & 2069 (29.19\%)\\ 
%  & \tactictoe (without special care for metis) & 154 & 3868 (\\
  & \tactictoe & 187 & 4186 (59.06\%) \\
%  & \tactictoe (+ self-learning) & 187 (187) &  4201 (59.27\%) \\
\bottomrule
\end{tabular}
\caption{Experiments with a timeout of 10 seconds \label{tab:_param}}
\end{table}

It is learning to use \metis however it cannot automatically learn to give it 
more time making it still a bit weaker than the adhoc startegy.

Table~\ref{theories} compares the success rates of re-proving for different
\holfour theories. \tactictoe(SH) outperforms \tactictoe(NH) on every 
considered theory.
Even if a lot weaker due to the missing premise selection component, 
\tactictoe(NH)
is hugely better than \holyhammer(blistr) in 4 theories: \texttt{measure}, 
\texttt{list}, \texttt{sorting} and \texttt{finite\_map}. The main reason is 
that 
those theories are equipped with specialized tactics, performing complex 
transformation such as structural induction and \tactictoe(NH) can reuse them. 
Conversely, \holyhammer(blistr) is more suited to deal with dense theories such 
as 
\texttt{real} or \texttt{complex} where a lot of related theorems are available 
and 
most proofs are usually completed by rewriting tactics.

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} 
& {meas} \\
\midrule
\tactictoe & 74.6 & 67.3 & 77.3 & 28.2\\
\eprover & 50.8 & 67.6 & 63.4 & 10.8\\
\midrule
\phantom{abc}  & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 37.3 & 71.4 & 58.4 & 73.6 \\
\eprover & 19.3 & 20.7 & 15.8 & 20.9 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories 
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},  
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}. 
}
\end{table}

\subsection{Time and space complexity}

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=south east, at={(0.9,0.1)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=4500,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=time, y=solved] {data/tt_ptime};
\addplot table[x=time, y=solved] {data/e_ptime};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Number of problem solved in less than x seconds.}
\end{figure}

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=size, y=solved] {data/tt_percent};
\addplot table[x=size, y=solved] {data/e_percent};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

%\begin{table*}[t]
%\centering\ra{1.3}
%\small
%\begin{tabular}{llllll}
%\toprule
% ID & \multicolumn{2}{c}{nodes}   & \multicolumn{2}{c}{proof size} & 
% \multicolumn{1}{c}{time}\\ \cmidrule(lr){2-3} \cmidrule(lr){4-5} 
% \cmidrule(lr){6-6}
%     & average & max & average & max & average \\
%\midrule
% \tactictoe & 94.66 & 421 & 3.34 & 39 & 0.66 \\
%\bottomrule
%\end{tabular}
% \caption{\label{tab:stats} old search statistics}
%\end{table*}

60 seconds experiment + 
adding theorems from the namespace and removing 
additionally saved theorems + 
recording proofs from Prove.

Constant progress: 
+45 percent in Septermber in 5 seconds.
+59 percent in December in 10 seconds.
+66.6 percent in January in 60 seconds.

The progress cannot be attributed only to the time increase but to testing and 
debugging of parameters.

>>>>

\subsection{The problem}\todo{}

Two ways of representing a formulas.

Defintion of a tactic in \holfour. In this research, we will take the definition
that a tactic takes a goal and returns a list of goals.

\subsection{Contributions}
This project is an extension of the work described in \cite{tgckju-lpar17}. We 
explain here the additional contributions and their effect across our system:

@@

%Mention explictly all the predictors that will be using this distance.
%tactics:
%theorems:
%terms:
%goal list: (product of tactics)
\begin{itemize}
\item tactic:\\
Given a goal $g$, the classifier selects a list of previously solved goals $l$
ordered by their similarity to $g$. Since each solved goal is associated 
with a tactic in our database, the ordered list of goals induces a ordered  
list of associated tactics. And tactics are tried in this order during the 
proof search. Therefore, to order the tactics correctly, it is crucial to 
estimate the similarity between goals with mathematically relevant features.
\item term:\\
Predict the closest term in a tactic that appears in the goal.

\item theorem:\\
Predict the theorem the closest to the current goal. Theorem dependencies are 
added during preselection.

\item list of goals:\\
Used for Monte Carlo Tree Search evaluation. Predict the closest list of goals
to the list of goal produced by a tactic.
\end{itemize}

@@

So, we first give a static description of the proof tree and its component.
Then we explain how we compute the evaluation and policy functions from learned 
data. And finally, we show how the two MCTS step, node selection and node 
extension changes the proof tree data structures.
Intuitively, MCTS offers a good trade-off between exploitation and exploration. 
It means that the algorithm will exploit deeper more promises branches but 
leaving enough time for exploring less likely alternatives.

@@

\begin{definition}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple 
$(\mathbb{A},\mathbb{G},\mathbb{T},A,G,T)$ 
that respects the following conditions:
\begin{itemize}
\item $\mathbb{A}$ is a set of nodes, $\mathbb{G}$ is a set of goals 
 and $\mathbb{T}$ is a set of tactics.
\item $A$ is a partial function from $\mathbb{G} \times \mathbb{T}$ to 
$\mathbb{A}$.
\item $T$ is a total function from $\mathbb{G}$ to $P(\mathbb{T})$.
\item $G$ is a total function from $\mathbb{A}$ to $P(\mathbb{G})$.
\item $Support(A) = \lbrace (g,t)\ |\ t \in T(g) \rbrace$.
\item if $t \in T(g)$ and $A(g,t) = a$ then $t(g) = G(a)$.
\end{itemize}

@@

In each node can either be solved, open or pending. The definition of a solved 
goal is expressed through mutual recursion with a 
solved node and is given in Definition~\ref{def:solved_goal}. An open goal is 
goal that 
was tried at least once by a tactic. In our implementation we have only one 
open goals per node at once. A goal that was not explored yet is called a 
pending goal.

@@

@@

Since the support of A induces a bijection between tactics of the tree 
search and the node they 
produced, any function define on tactics of the tree search can be map to a 
function defined on nodes. 
Therefore, we define $PriorPolicy(a)$ as the $PriorPolicy(t)$ where $t$ is the 
tactic that produced $a$.

@@

@@

%Mention the conflicts.
%Theorem orthogonalization is not turn on during instantiation of tacitcs 
%except 
%for the metis tactic.
%Best parameters + comparison with holyhammer.

\eprover is not called by \tactictoe for fair comparison.

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{lll}
\toprule
  & & 7164 theorems \\
\midrule
  & \eprover   & 2472 (34.5\%)\\ 
  & \tactictoe & 4741 (66.2\%)\\
  & combined   & 4946 (69.0\%)\\
\bottomrule
\end{tabular}
\caption{Experiments with a timeout of 60 seconds \label{tab:_param}}
\end{table}

Table~\ref{theories} compares the success rates of re-proving for different
\holfour theories. \tactictoe outperforms \eprover on every 
considered theory.
\eprover is more suited to deal with dense theories such as 
\texttt{real} or \texttt{complex} where a lot of related theorems are available 
and most proofs are usually completed by rewriting tactics.

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} & {meas} \\
\midrule
\tactictoe & 81.2 & 74.0 & 79.6 & 31.3\\
\eprover & 59.9 & 72.0 & 67.1 & 12.8\\
\midrule
\phantom{abc} & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 45.8 & 79.5 & 65.3 & 82.0 \\
\eprover & 24.1 & 26.5 & 15.8 & 24.7 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories 
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},  
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}. 
}
\end{table}

\subsection{Slight improvements}
The final results could be improved by a few percent by including the 
following additional techniques that are already available.

\paragraph{Asynchronous calls to Eprover inside TacticToe}
The technique already discussed in Section~ would most likely improve the 
results by 2-3 percent.

\paragraph{Self-learning}
Until now, all our feature vectors were learned form human proofs. 
human proofs. We can also add to this tactics that were part of a proof found 
by our algorithm. To prevent duplication of effort, orthogonalization of those 
tactics is essential to have a beneficial effect (see Experiments).
Since recording and re-proving are intertwined, the 
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own 
discovered proofs than from the human proof 
scripts~\cite{DBLP:conf/cade/Urban07}. It is also the first step toward 
self-reinforcement learning. Typically this technique improves the results by 
less than a percent.

\subsection{Complexity of the proofs}

Estimate the difficulty of the proof by the number of tactic units.

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=2000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}     
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe,original}
\end{axis}
\end{tikzpicture}
\caption{Distribution of the proof length in tactic unit of evaluated theorems.}
\end{figure}

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=south east, at={(0.9,0.1)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=60,
  ymin=0, ymax=4800,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=time, y=solved] {data/tactictoe_time};
\addplot table[x=time, y=solved] {data/eprover_time};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Number of problem solved in less than x seconds.}
\end{figure}
\eprover uses at least two strategies where as \tactictoe only uses one set 
of parameters.

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=20,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=oplen, y=solved] {data/tactictoe_by_oplen};
\addplot table[x=oplen, y=solved] {data/eprover_by_oplen};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

%60 seconds experiment + 
%adding theorems from the namespace and removing 
%additionally saved theorems + 
%recording proofs from Prove.

% 
%The progress cannot be attributed only to the time increase but to testing and 
%debugging of parameters.
